---
title: "Models for Football Data"
output:
  html_document:
    df_print: paged
    code_folding: hide
---
# Data Processing
We load football data from a Kaggle containing Spanish League results, for the 2016-17 and 2017-18 seasons.

```{r message=FALSE, warning=FALSE}
library("dplyr")
library("tidyr")
library("ggplot2")
library("reshape2")
library("nloptr")
library("data.table")
data <- read.csv("FMEL_Dataset.csv") 

data <- data %>%
  filter(season == "2016-17" | season == "2017-18" )

head(data)
```

Reshaping the data so that we have one team per row.
```{r}
melt_results <- function(results_df) {
  results_df %>%
    # select only relevant columns
    select(localTeam, visitorTeam, localGoals, visitorGoals, date) %>%
    gather(location, team,  -localGoals, -visitorGoals, -date) %>%
    # calculate goals for/against the team
    mutate(g_for = case_when(
      location == "localTeam" ~ localGoals,
      location == "visitorTeam" ~ visitorGoals
    )) %>%
    mutate(g_ag = case_when(
      location == "localTeam" ~ visitorGoals,
      location == "visitorTeam" ~ localGoals
    )) 
}

data_melted <- data %>%
  melt_results()

head(data_melted)
```


Creating a ranking table for each team's statistics and ranking estimates. W,L and D are each team's number of wins, loses and draws respectively. GF,GA and GD represent the sum of goals for, against and the goal differences for each team. Points are calculated by the formula “Pts = 3W + 1D” and teams are ranked accordingly. 
```{r}
game_statistics <- data_melted %>%
  mutate(win = ifelse(g_for > g_ag, 1,0),
         loss = ifelse(g_for < g_ag, 1,0),
         draw = ifelse(g_for == g_ag, 1,0))

game_statistics <- game_statistics %>%
  group_by(team) %>%
  summarise ( W = sum(win),
           L = sum (loss),
           D = sum(draw),
           GF = sum(g_for),
           GA = sum(g_ag),
           GD = GF - GA,
           Pts = 3*W + 1*D
           ) %>%
  arrange(desc(Pts)) 
  
head(game_statistics)
```

# Poisson Model - 1 (Maher)

We group all teams together into “home” and “away” categories  and plot the distribution of the goals.

```{r}
data_melted %>%
  ggplot(., aes(x = g_for, fill = location)) +
  geom_density(adjust = 8, alpha = 0.5) +
  scale_fill_manual(values = c("green", "red")) +
  scale_x_continuous(breaks = 0:6) +
  labs(title = "Goals scored at home and away",
       x = "goals scored",
       y = "density") +
  theme_minimal()
```

It is evident that the home team tends to score more than the visiting one. Also both charts look like a Poisson distribution.For the home team we take the average number of goals scored per match as the mean of a Poisson distribution and we can see how well the previous graph fits a simulated Poisson process.


```{r}
p1 <- data_melted %>%
  filter( location == "localTeam") %>%
  ggplot(., aes(x = g_for)) +
  geom_density(adjust = 8, alpha = 0.5, aes(fill= "green")) +
  scale_fill_manual(values = c("green")) +
  scale_x_continuous(breaks = 0:6) +
  labs(title = "Goals scored at home vs simulated Poisson distribution",
       x = "goals scored",
       y = "density") +
  theme_minimal() + 
  theme(legend.position = "none")

local_team_data <- data_melted %>%
  filter( location == "localTeam")

simulated_poisson <- data.frame(dist = rpois(100000, mean(local_team_data$localGoals)))

p1 + geom_density(data = simulated_poisson, aes(x = dist),
                  fill = NA, adjust = 8, alpha = 0.2) 
```
In a similar fashion one can conclude that the goals scored by the visiting team also follow a Poisson process but with a different mean. 

Hence, the probability of team i scoring x goals against team j is:
$$
P(x_i=x) = \frac{\alpha^x_ie^{−\alpha_i}}{x!} 
$$
Let n be the number of teams and $x_k$ and $y_k \in 0,1,2,3,..$ are the final scores of the home and away team, respectively in the $k$-th match.

Where $\alpha$ can be approximated using the maximum likelihood estimator for Poisson distribution, namely

$$
\hat{\alpha}_i= \frac{1}{m_i}\sum_{k=1}^{m_i} x_k
$$
and $m_i$ is the number of matches that team $i$ participated in. Essentially $\alpha$ is an estimate of each team's attack rating. Similarly, we can estimate each team's defence parameter $\beta_j$ as the average number of goals scored by their opponents which also  follows a Poisson process. 

Lastly we introduce $\gamma$ as a measure of the "home advantage". As we saw earlier, in general football teams tend to perform better at home, so we want to factor that in. 

$\gamma$ can be calculated as the sum of home goals divided by the sum of away goals. 

$$
\gamma = \frac{\sum x}{\sum y}
$$
Ultimately we arrive at the following Poisson model which takes into account the team's attack, the defence of its opponent and if the team is playing at home. 

$$
X_i,_j \sim Poisson(\alpha_i\beta_j\gamma)
$$
$$
Y_i,_j\sim Poisson(\alpha_j\beta_i)
$$
Estimating gamma - home advantage
```{r}
gama = sum(data$localGoals)/sum(data$visitorGoals)
gama
```

Estimating alfa and beta - average number of goals for and against each team

```{r}
data_model1 <- data_melted %>%
  group_by(team) %>%
  summarise(alfa = mean(g_for),
            beta = mean(g_ag))

head(data_model1)
```

The following code estimates the outcome for each game in our dataset based on the calculated parameters. The local team estimate is  multiplied by gamma to account for the "home advantage".

```{r}
model1_output <- left_join(data, data_model1[c("team", "alfa", "beta")], by = c("localTeam" = "team"))
model1_output <- model1_output %>%
  rename(alfa_local=alfa) %>%
  rename(beta_local=beta)

model1_output <- left_join(model1_output, data_model1[c("team", "alfa", "beta")], by = c("visitorTeam" = "team"))
model1_output <- model1_output %>%
  rename(alfa_visitor=alfa) %>%
  rename(beta_visitor=beta)

model1_output <- model1_output %>% 
  mutate(prediction_local = alfa_local*beta_visitor*gama,
         prediction_visit = alfa_visitor*beta_local)

head(model1_output)
```

We check that the estimated parameters indeed predict the observed results for each team.The top three teams predicted by the model are indeed in top 3 of the real ranking but in different order. Overall the ranking of the top six teams according to the model does not match real life results.  

```{r}
game_predicted_statistics <- model1_output %>%
  # select only relevant columns
  select(localTeam, visitorTeam, prediction_local, prediction_visit, date) %>%
  gather(location, team,  -prediction_local,-prediction_visit, -date) %>%
  # calculate goals for/against the team
  mutate(g_for = case_when(
    location == "localTeam" ~ prediction_local,
    location == "visitorTeam" ~ prediction_visit
  )) %>%
  mutate(g_ag = case_when(
    location == "localTeam" ~ prediction_visit,
    location == "visitorTeam" ~ prediction_local
  )) %>%
  mutate(win = ifelse(g_for > g_ag, 1,0),
         loss = ifelse(g_for < g_ag, 1,0),
         draw = ifelse(g_for == g_ag, 1,0)) %>%
  group_by(team) %>%
  summarise ( W = sum(win),
              L = sum (loss),
              D = sum(draw),
              GF = sum(g_for),
              GA = sum(g_ag),
              GD = GF - GA,
              Pts = 3*W + 1*D
  ) %>%
  arrange(desc(Pts))

head(game_predicted_statistics)
  
```
# Alternative approach to solving for the parameters of Model 1

Once again we are interested in the following model which assumes independence of $X_i,_j$ and $Y_i,_j$:

$$
P(X_i,_j = x, Y_i,_j = y) = \frac{\lambda e^{-\lambda}}{x!}\frac{\mu e^{-\mu}}{x!}
$$
where

$$
\lambda = \alpha_i\beta_j\gamma
$$
$$
\mu = \alpha_j\beta_i
$$
The likelihood function of the Poisson model is given by

$$
L(\alpha_i,\beta_i,\gamma,i = 1,2,...,n) = \prod_{k=1}^{N} e^{-\lambda_k}\lambda_k^{x_k} e^{-\mu_k}\mu_k^{x_k} 
$$
We'd like to find the set of parameters that will minimize the likelihood function.


We start by creating a new data set containing only the relevant information.

```{r}
x = data[,c(5:8)]
n = n_distinct(x$localTeam)
teams <- matrix(c(unique(x$localTeam),1:n), nrow = n)
teams <- as.data.frame(teams)
x <- left_join(x,teams, by = c("localTeam" = "V1"))
x <- left_join(x,teams, by = c("visitorTeam" = "V1"))

x <- x %>%
  rename(local_team_index = V2.x,
         visitor_team_index = V2.y)
x <-x [,c(3:6)]

x <- as.data.table(x)

x$local_team_index <- as.integer(x$local_team_index)
x$visitor_team_index <- as.integer(x$visitor_team_index)

x <- as.data.table(x)
head(x)
```


In what follows we apply a constraint optimization algorithm where we are using the following constraint:

$$
\sum_{i=1}^n\frac{\alpha_i}{n}=1
$$
or the equivalent 

$$
ln(\sum_{i=1}^n\frac{\alpha_i}{n}=1) = 0
$$
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
f_lambda = function(alfa, beta, gama){
  gama*alfa[x$local_team_index]*beta[x$visitor_team_index] 
}

f_mu = function(alfa, beta, gama){
  alfa[x$local_team_index]*beta[x$visitor_team_index]
}

f_l = function(w) {
  alfa = w[1:n] 
  beta = w[(n+1):(2*n)]
  gama = w[2*n+1]
  lambda = f_lambda(alfa, beta, gama)
  mu =     f_mu(alfa, beta, gama)
  
  - sum(x$localGoals * log(lambda) + x$visitorGoals * log(mu) - lambda - mu) 
}

eq = function(w) {
  alfa = w[1:n]
  c(log(sum(alfa)/n))
}

w0 = rep(1, 2*n+1)
options(warn = -1)

suppressMessages(
S <- slsqp(w0, fn = f_l,
           heq = eq,
           control = list(xtol_rel = 1e-9, print_level = 0)
            )
)
S
```

We rank teams based on the combination of their attack and defense.

```{r}
alpha <- S$par[1:n]
beta <- S$par[(n+1):(2*n)]
teams$alpha <- alpha
teams$beta <- beta
names(teams)[1]<-"team"
names(teams)[2]<-"team index"
game_statistics_mod1 <- left_join(game_statistics,teams, by = c("team"))
game_statistics_mod1 <- game_statistics_mod1 %>%
  mutate(skill = alpha*beta) %>%
  arrange(desc(skill))

head(game_statistics_mod1)
```

Note that some of the teams with low real life ranking like Betis (30th position in the ral ranking) have pretty high attack and defense scores.

# Poisson Model 2 (Dixon-Coles)
The Dixon-Coles model is very similar to the one described in the previous section. They relaxed the assumption of independence between the scores of the two teams by introducing the following function:

$$
\tau_\lambda,_\mu(x,y) = \begin{cases}
            1-\lambda\mu\rho \qquad \qquad\ \text{if } x=y=0\\
            1+\lambda\rho     \qquad\qquad \text{if } x=0, y=1\\
            1+\mu\rho         \quad \quad\quad\ \ \ \ \text{if } x=1, y=0\\
            1-\rho            \qquad \quad \quad\quad\ \ \text{if } x=y=1\\
            1                  \qquad \qquad \qquad \qquad \ \text{otherwise}
            \end{cases}
$$
and proposed the enhanced model:
$$
P(X_i,_j = x, Y_i,_j = y) = \tau_\lambda,_\mu(x,y)\frac{\lambda e^{-\lambda}}{x!}\frac{\mu e^{-\mu}}{x!}
$$

The likelihood function of the model is given by

$$
L(\alpha_i,\beta_i,\gamma,\rho,i = 1,2,...,n) = \prod_{k=1}^{N} \tau_{\lambda_k},_{\mu_k}(x,y)e^{-\lambda_k}\lambda_k^{x_k} e^{-\mu_k}\mu_k^{x_k} 
$$
In what follows we apply the same constraint optimization algorithm as before but ingluding the new parameter rho.

```{r message=FALSE, warning=FALSE}
f_lambda = function(alfa, beta, gama){
  gama*alfa[x$local_team_index]*beta[x$visitor_team_index] 
}

f_mu = function(alfa, beta, gama){
  alfa[x$local_team_index]*beta[x$visitor_team_index]
}

tau<- rep(0,nrow(x))

f_tau = function(lambda, mu, rho){
  for(i in 1:nrow(x)){
  if (x$localGoals[i]== 0 && x$visitorGoals[i] == 0) {
      tau[i] = 1-lambda[i]*mu[i]*rho
    } else if (x$localGoals[i]== 0 && x$visitorGoals[i] == 1) {
      tau[i] = 1+lambda[i]*rho
    } else if (x$localGoals[i]== 1 && x$visitorGoals[i] == 0) {
      tau[i] = 1+mu[i]*rho
    } else if (x$localGoals[i] == 1 && x$visitorGoals[i] == 1) {
      tau[i] = 1-rho
    } else {tau[i] = 1}
  }
  return(tau)}


f_l = function(w) {
  alfa = w[1:n] 
  beta = w[(n+1):(2*n)]
  gama = w[2*n+1]
  rho = w[2*n+2]
  lambda = f_lambda(alfa, beta, gama)
  mu = f_mu(alfa, beta, gama)
  tau = f_tau(lambda, mu, rho)
  
  - sum(x$localGoals * log(lambda) + x$visitorGoals * log(mu) - lambda - mu + log(tau)) 
}

eq = function(w) {
  alfa = w[1:n]
  c(log(sum(alfa)/n))
}

w0 = c(rep(1, 2*n+1),0)
options(warn = -1)

suppressMessages(
S_DC <- slsqp(w0, fn = f_l,
           heq = eq,
           control = list(xtol_rel = 1e-9, print_level = 0)
            )
)
S_DC
```


Ranking the teams based on the newly calculated attack and defense parameters (rho and the home effect are not taken into account).
```{r}
home_adv <- S_DC$par[2*n+1]
rho_par <- S_DC$par[2*n+2]
alpha <- S$par[1:n]
beta <- S$par[(n+1):(2*n)]
teams$alpha <- alpha
teams$beta <- beta
game_statistics_mod2 <- left_join(game_statistics,teams, by = c("team"))
game_statistics_mod2 <- game_statistics_mod2 %>%
  mutate(skill = alpha*beta) %>%
  arrange(desc(skill))
head(game_statistics_mod2)
```

Calculating the probability of the possible results (0-0 , 0-1, 1-0, etc.) for each game using the model. 


```{r}
data_combined <- left_join(data, teams[c("team", "alpha","beta")], by = c("localTeam" = "team"))
data_combined <- left_join(data_combined, teams[c("team","alpha","beta")], by = c("visitorTeam" = "team"))
data_combined <- data_combined %>%
 rename(alpha_local = alpha.x,
        beta_local = beta.x,
        alpha_visitor = alpha.y,
        beta_visitor = beta.y) %>%
  mutate(lambda= alpha_local*beta_visitor*home_adv,
         mu = alpha_visitor*beta_local,
         pred_0_0 = (1-lambda*mu*rho_par)*exp(-lambda)*(lambda)^0*exp(-mu)*(mu)^0/(factorial(0)*factorial(0)),
         pred_1_0 = (1+mu*rho_par)*exp(-lambda)*(lambda)^1*exp(-mu)*(mu)^0/(factorial(1)*factorial(0)),
         pred_0_1 = (1+lambda*rho_par)*exp(-lambda)*(lambda)^0*exp(-mu)*(mu)^1/(factorial(0)*factorial(1)),
         pred_1_1 = (1-rho_par)*exp(-lambda)*(lambda)^1*exp(-mu)*(mu)^1/(factorial(1)*factorial(1)),
         pred_2_0 = exp(-lambda)*(lambda)^2*exp(-mu)*(mu)^0/(factorial(2)*factorial(0)),
         pred_0_2 = exp(-lambda)*(lambda)^0*exp(-mu)*(mu)^2/(factorial(0)*factorial(2)),
         pred_2_1 = exp(-lambda)*(lambda)^2*exp(-mu)*(mu)^1/(factorial(2)*factorial(1)),
         pred_1_2 = exp(-lambda)*(lambda)^1*exp(-mu)*(mu)^2/(factorial(1)*factorial(2)),
         pred_2_2 = exp(-lambda)*(lambda)^2*exp(-mu)*(mu)^2/(factorial(2)*factorial(2)))

options("scipen"=100, "digits"=2)
head(data_combined[,c("localTeam", "visitorTeam", "localGoals", "visitorGoals","pred_0_0","pred_1_0","pred_0_1","pred_1_1","pred_2_0","pred_0_2","pred_2_1","pred_1_2","pred_2_2")])
```



